---
title: "DATA 607 Final Project"
author: "Samuel I Kigamba"
date: "December 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project Checklist
To receive full credit, youâ????ll need to deliver on all of the items in the checklist below. Please read carefully through this checklist
before you make your project proposal. You are (within these checklist constraints) strongly urged to limit scope and make the
necessary simplifying assumptions so that you can deliver your work on time!

# Section 1 - Introduction


##ï¯ Proposal describes your motivation for performing this analysis.

Goal:
My objective in this project is seeking to understand the relationship between population growth and the crime rates in the United States between the year 1994 to 2013 and to comment and conclude on the overal pattern.


#ï¯ Proposal describes likely data sources.

 To analyze the crime rates in the US for the period from 1994 to 2013, we will use the data set provided by the FBI. 
 See link: https://ucr.fbi.gov/crime-in-the-u.s/2013/crime-in-the-u.s.-2013/tables/1tabledatadecoverviewpdf/table_1_crime_in_the_united_states_by_volume_and_rate_per_100000_inhabitants_1994-2013.xls
 
 

#ï¯ Your project has a recognizable â????data science workflow,
â?? such as the OSEMN workflow or Hadley Wickhamâ????s Grammar of Data Science. [Example: First the data is acquired, then necessary transformations and clean-up are performed, then the analysis and presentation work is performed]

## Data Acquisition
I downloaded data for the period 1994 - 2018 and saved it as a .csv file in my local drive and later uploaded it into my github directory for universal accessibility.

load data

```{r}
crime_data <- read_csv("CrimeData.csv")
head(crime_data)

```


#Data Transformation and Clean-up

```{r}
crime_data <- crime_data[3:28, 1:20]
 
head(crime_data)

colNames <- crime_data[1,]
colNames

colnames(crime_data) <- colNames
head(crime_data)
 
crime_data <- crime_data[-1,]
 
crime_data <- as_tibble(crime_data)
 
head(crime_data)

```


#ï¯ Project includes data from at least two different types of data sources
(e.g., two or more of these: relational or CSV, Neo4J, web page [scraped or API], MongoDB, etc.)


#ï¯ Project includes at least one data transformation operation.
[Examples: transforming from wide to long; converting columns to date format]


#ï¯ Project includes at least one statistical analysis and at least one graphics that describes or validates your data.

```{r}
populationGrowth <-
        ggplot(crime_data, aes(Year, Population, fill = Population)) + 
        geom_bar(stat = "identity") +
        ggtitle("Population growth between 1994 - 2018") +
        xlab("Years") + ylab("Population") +
        theme(
                plot.title = element_text(color="blue", size=15, face="bold.italic"),
                axis.text.x = element_text(angle=60, hjust=1),
                #axis.text.y = element_blank(),
                #axis.ticks.y = element_blank(),
                axis.title.x = element_text(color="blue", size=15, face="bold"),
                axis.title.y = element_text(color="blue", size=15, face="bold")
                )
 
ggplotly(populationGrowth)

```

```{r}
ViolentCrimeRate <-
   ggplot(crime_data, aes(Year, Violent_crime, fill = Violent_crime)) + 
   geom_bar(stat = "identity") +
   ggtitle("Violent_crime growth between 1994 - 2018") +
   xlab("Years") + ylab("Violent_crime") +
   theme(
     plot.title = element_text(color="blue", size=15, face="bold.italic"),
     axis.text.x = element_text(angle=60, hjust=1),
     #axis.text.y = element_blank(),
     #axis.ticks.y = element_blank(),
     axis.title.x = element_text(color="blue", size=15, face="bold"),
     axis.title.y = element_text(color="blue", size=15, face="bold")
   )
 
ggplotly(ViolentCrimeRate)
 
names(crime_data)

```
```{r}

crime_data %>% mutate(OverallCrimeRate = (
   Violent_crime_rate + Murder_and_nonnegligent_manslaughter_rate +
   Rape_rate + Robbery_rate + Aggravated_assault_rate +
   Property_crime_rate + Burglary_rate + Larceny-theft_rate +
   Motor_vehicle_theft_rate) / 9
 ) -> crimeData

```

#ï¯ Project includes at least one graphic that supports your conclusion(s).


#ï¯ Project includes at least one statistical analysis that supports your conclusion(s).









#ï¯ Project includes at least one feature that we did not cover in class!
        There are many examples: â????I used ggmap; I created a decision tree; I ranked the results; I created my presentation slides directly from R; I figured out to use OAuth 2.0â??¦â??
        
#ï¯ Presentation. Was the presentation delivered in the allotted time (3 to 5 minutes)?

#ï¯ Presentation.
Did you show (at least) one challenge you encountered in code and/or data, and what you did when you encountered that challenge? If you didnâ????t encounter any challenges, your assignment was clearly too easy for you!

#ï¯ Presentation.
Did the audience come away with a clear understanding of your motivation for undertaking the project?

#ï¯ Presentation.
Did the audience come away with a clear understanding of at least one insight you gained or conclusion you reached or hypothesis you â????confirmedâ?? (rejected or failed to rejectâ??¦)?

#ï¯ Code and data.
Have you delivered the submitted code and data where it is self-containedâ????preferably in rpubs.com and github? 
Am I able to fully reproduce your results with what youâ????ve delivered? You wonâ????t receive full credit if your code references data on your local machine!

#ï¯ Code and data.
Does all of the delivered code run without errors?

#ï¯ Code and data.
Have you delivered your code and conclusions using a â????reproducible researchâ?? tool such as RMarkdown?

#ï¯ Deadline management.
Were your draft project proposal, project, and presentation delivered on time?
Any part of the project that is turned in late will receive a maximum grade of 80%. Please turn in your work on time!
You are of course welcome to deliver ahead of schedule!